---
title: "Bayesian Instrumental Variable analysis for Binary Outcomes"
author: "Yichi Zhang"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Below we demonstrate the overall workflow of the BIVA approach for binary outcomes. 

```{r start, results='hide', message=FALSE, warning=FALSE}
library(biva)
```

## Demonstration: two-side noncompliance

### Data simulation

We simulate some data from a hypothetical randomized experiment 
with noncompliance. Those assigned to the treatment group can
decide to use the feature or not. Those assigned to the control group 
may also use the feature since it is available to all. In this case,
we have two-side noncompliance where the assigned treated units can opt out
and the assigned control can opt in. The outcome metric is a binary variable.

Therefore, there are 3 principal strata: compliers, never-takers, and
always-takers.

```{r simulated_data_two_side}
set.seed(1997)
n <- 200

# Covariates: X observed, U unobserved
X <- rnorm(n)
U <- rbinom(n, 1, 0.5)

## True memberships of principal strata (1:c,2:nt,3:at): S-model depends only on U
true.PS <- rep(0, n)
U1.ind <- (U == 1)
U0.ind <- (U == 0)
num.U1 <- sum(U1.ind)
num.U0 <- sum(U0.ind)
true.PS[U1.ind] <- t(rmultinom(num.U1, 1, c(0.6, 0.3, 0.1))) %*% c(1, 2, 3)
true.PS[U0.ind] <- t(rmultinom(num.U0, 1, c(0.4, 0.5, 0.1))) %*% c(1, 2, 3)

## Treatment assigned: half control & half treatment
Z <- c(rep(0, n / 2), rep(1, n / 2))

## Treatment received: determined by principal strata and treatment assigned
D <- rep(0, n)
c.trt.ind <- (true.PS == 1) & (Z == 1)
c.ctrl.ind <- (true.PS == 1) & (Z == 0)
nt.ind <- (true.PS == 2)
at.ind <- (true.PS == 3)
num.c.trt <- sum(c.trt.ind)
num.c.ctrl <- sum(c.ctrl.ind)
num.nt <- sum(nt.ind)
num.at <- sum(at.ind)
D[at.ind] <- rep(1, num.at)
D[c.trt.ind] <- rep(1, num.c.trt)

## Generate observed outcome: Y-model depend on X, U, D, and principal strata
Y <- rep(0, n)
Y[c.ctrl.ind] <- rbinom(num.c.ctrl, 1, 0.3 + 0.02 * X[c.ctrl.ind] - 0.2 * U[c.ctrl.ind])
Y[c.trt.ind] <- rbinom(num.c.trt, 1, 0.5 + 0.02 * X[c.trt.ind] - 0.2 * U[c.trt.ind])
Y[nt.ind] <- rbinom(num.c.trt, 1, 0.7 + 0.02 * X[nt.ind] - 0.2 * U[nt.ind])
Y[at.ind] <- rbinom(num.c.trt, 1, 0.4 + 0.02 * X[at.ind] - 0.2 * U[at.ind])

df <- data.frame(Y = Y, Z = Z, D = D, X = X, U = U)
```

For the strata model, when U = 1, a unit is a complier, a never-taker, 
or an always-taker with probability 0.6, 0.3, and 0.1, respectively.
When U = 0, a unit is a complier, a never-taker, or an always-taker with
probability 0.4, 0.5, and 0.1, respectively.

The outcome models are by different strata-nudge combinations.
The compliers may adopt the treatment or not depending on the assigned nudge, 
and thus have two outcome models by treatment groups. The never-takers will 
never adopt the treatment feature. The always-takers will always adopt the
treatment feature. We assume away the direct effect of 
the nudge here so there is only one outcome model for never-takers, and
also one model for always-takers.

The four outcome models depend on both the observed X and the unobserved U
with the same slopes. The contrast between the intercepts of 
the outcome models for compliers then becomes the CACE of the feature, 
which is 0.5 - 0.3 = 20%.

### OLS analysis

We again fit an OLS model with the adopted treatment D and 
the observed covariate X to see the caveats.

```{r OLS_two_side}
OLS <- lm(data = df, formula = Y ~ D + X)
summary(OLS)
confint(OLS, "D")
```

It returns an uninformative insignificant negative estimate of the treatment
effect.

We restart the BIVA workflow to get more insight into the impact.

### Prior predictive checking

The outcome type is specified by having y_type = "binary".
We assume exclusion restriction, so ER = 1. 
With two-side noncompliance, we set side = 2.
The parameters in the prior distributions are specified as well.
We first look into what the prior distributions assume about the parameters.

```{r priors_check_two_side, fig.align = 'center'}
ivobj <- biva$new(
  data = df, y = "Y", d = "D", z = "Z",
  x_ymodel = c("X"),
  x_smodel = c("X"),
  y_type = "binary",
  ER = 1,
  side = 2,
  beta_mean_ymodel = matrix(0, 4, 2),
  beta_sd_ymodel = matrix(1, 4, 2),
  beta_mean_smodel = matrix(0, 2, 2),
  beta_sd_smodel = matrix(1, 2, 2),
  fit = FALSE
)

ivobj$plotPrior()
```

By prior, the effect of the AI-powered dubbing feature among compliers 
is negligible.
The proportions of compliers, never-takers, and always-takers are similar.

### Fitting a BIVA model to the data

Now we fit a BIVA model to the data.

```{r fit_two_side, fig.align = 'center'}
ivobj <- biva$new(
  data = df, y = "Y", d = "D", z = "Z",
  x_ymodel = c("X"),
  x_smodel = c("X"),
  y_type = "binary",
  ER = 1,
  side = 2,
  beta_mean_ymodel = matrix(0, 4, 2),
  beta_sd_ymodel = matrix(1, 4, 2),
  beta_mean_smodel = matrix(0, 2, 2),
  beta_sd_smodel = matrix(1, 2, 2),
  fit = TRUE
)
```

We look at the trace plot of outcomes in each strata.
The four plots are the posterior draws of the mean outcomes among 
the compliers assinged to control, the compliers nudged to treatment,
the never-takers, and the always-takers.

```{r tracePlot_two_side, fig.align = 'center'}
ivobj$tracePlot()
```

The convergence and mixing look good.

We run the weak instrument test.

```{r weak_IV_test_two_side, fig.align = 'center'}
ivobj$weakIVTest()
```

No weak instrument issue is detected.

We can use the following methods to summarize our findings:

```{r findings_two_side, fig.align = 'center'}
# Posterior distribution of the strata probability
ivobj$strataProbEstimate()
# Posterior probability that CACE is greater than 0.1
ivobj$calcProb(a = 0.1)
# Posterior median of the CACE
ivobj$pointEstimate()
# Posterior mean of the CACE
ivobj$pointEstimate(median = FALSE)
# 75% credible interval of the CACE
ivobj$credibleInterval()
# 95% credible interval of the CACE
ivobj$credibleInterval(width = 0.95)
```
The posterior returns informative conclusions on the impact of the treatment.

Visualizations of how the data updates our knowledge about the impact
by comparing its posterior distribution with the prior can be done by
running the following methods. We hide them here.

```{r comparison_two_side, eval = FALSE}
ivobj$vizdraws(
  display_mode_name = TRUE, breaks = 0.1,
  break_names = c("< 0.1", "> 0.1")
)
ivobj$lollipop(threshold = 0.1, mediumText = 8)
```

---

#### BIVA

Licensed under the Apache License, Version 2.0.\
