---
title: "Bayesian Instrumental Variable analysis for Count Outcomes"
author: "Yichi Zhang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Instrumental Variable analysis for Count Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Motivation

### The Challenge with Count Outcomes

In many experimental settings, the outcome of interest is a **count variable** rather than a continuous measure.

Standard Instrumental Variable (IV) methods (like 2SLS) often treat outcomes as continuous. When applied to count data, linear models can be problematic because:

1.  **Impossible Predictions:** They may predict negative counts.
2.  **Skewness:** Count data is often highly right-skewed with many zeros.
3.  **Mean-Variance Relationship:** Unlike Gaussian data, the variance of count data often increases with the mean (heteroscedasticity).

### The BIVA Solution

The **BIVA** package supports `y_type = "count"` to handle these cases explicitly using a **Negative Binomial** model. This approach:

  * Uses a **log-link** function to ensure predictions are always non-negative.
  * Estimates a **dispersion parameter ($\phi$)** to account for over-dispersion (where Variance \> Mean), which is common in real-world data and often violates simple Poisson assumptions.
  * Calculates the **CACE** (Complier Average Causal Effect) on the interpretable **natural count scale** (difference in expected counts), not just the log scale.

-----

## Workflow Demonstration

```{r start, message=FALSE, warning=FALSE}
library(BIVA)
library(ggplot2)
library(MASS) # For naive GLM comparison
```

### Data Simulation

We simulate data for a hypothetical experiment:

  * **Z (Instrument):** A randomized email "nudge" sent to users.
  * **D (Treatment):** Whether the user actually used the new feature.
  * **Y (Outcome):** The number of times the user logged in over the next month (Count).
  * **Compliance:** Two-sided noncompliance (Compliers, Never-takers, Always-takers).

We generate the data using a Negative Binomial process where the treatment increases the expected count of logins for Compliers.

```{r data_sim}
set.seed(2025)
n <- 1000

# Covariates: X (Observed), U (Unobserved Confounder)
X <- rnorm(n)
U <- rbinom(n, 1, 0.5)

# S-model: Principal Strata (1:Complier, 2:Never-taker, 3:Always-taker)
# U affects compliance (confounding)
true.PS <- rep(0, n)
U1.ind <- (U == 1); U0.ind <- (U == 0)

# U=1 increases prob of being a Complier
true.PS[U1.ind] <- t(rmultinom(sum(U1.ind), 1, c(0.50, 0.40, 0.10))) %*% c(1, 2, 3)
# U=0 decreases prob of being a Complier
true.PS[U0.ind] <- t(rmultinom(sum(U0.ind), 1, c(0.30, 0.60, 0.10))) %*% c(1, 2, 3)

# Treatment Assignment (Z) & Receipt (D)
Z <- c(rep(0, n/2), rep(1, n/2))
D <- rep(0, n)

# Strata definitions
c.trt.ind <- (true.PS == 1) & (Z == 1)
c.ctrl.ind <- (true.PS == 1) & (Z == 0)
nt.ind <- (true.PS == 2)
at.ind <- (true.PS == 3)

# D is 1 for Treated Compliers and Always-takers
D[c.trt.ind] <- 1
D[at.ind] <- 1

# Y-model: Negative Binomial Outcome
# Log link: log(mu) = intercept + beta*X + gamma*U
# True Dispersion phi = 5
phi_true <- 5
Y <- rep(0, n)

# Compliers (Control): Baseline ~ exp(2.0) = 7.4 logins
Y[c.ctrl.ind] <- rnbinom(sum(c.ctrl.ind), 
                         mu = exp(2.0 + 0.3*X[c.ctrl.ind] - 0.5*U[c.ctrl.ind]), 
                         size = phi_true)

# Compliers (Treated): Effect ~ exp(2.5) = 12.2 logins
# True CACE on Count Scale ~ 12.2 - 7.4 = 4.8 additional logins
Y[c.trt.ind] <- rnbinom(sum(c.trt.ind), 
                        mu = exp(2.5 + 0.3*X[c.trt.ind] - 0.5*U[c.trt.ind]), 
                        size = phi_true)

# Never-takers: High baseline ~ exp(2.8) = 16.4 logins
Y[nt.ind] <- rnbinom(sum(nt.ind), 
                     mu = exp(2.8 + 0.3*X[nt.ind] - 0.5*U[nt.ind]), 
                     size = phi_true)

# Always-takers: Low baseline ~ exp(1.5) = 4.5 logins
Y[at.ind] <- rnbinom(sum(at.ind), 
                     mu = exp(1.5 + 0.3*X[at.ind] - 0.5*U[at.ind]), 
                     size = phi_true)

df <- data.frame(Y = Y, Z = Z, D = D, X = X, U = U)

# Visualize the count distribution
ggplot(df, aes(x = Y, fill = factor(D))) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "identity") +
  labs(title = "Distribution of Login Counts by Treatment Received", x = "Logins", fill = "Treatment (D)") +
  theme_minimal()
```

### Naive Analysis (Biased)

First, we attempt a standard generalized linear model (GLM) using Negative Binomial regression, treating the received treatment $D$ as exogenous. This ignores the unobserved confounder $U$.

```{r naive_glm}
# Naive Negative Binomial Regression (As-Treated)
naive_model <- MASS::glm.nb(Y ~ D + X, data = df)
summary(naive_model)$coefficients["D",,drop=FALSE]
```

The naive estimate is likely biased because $D$ is correlated with $U$ (confounding). In our simulation, users who are "Always-takers" (who take D=1) have systematically lower baseline outcomes, pulling the naive estimate down.

### BIVA Analysis

We now use BIVA to recover the causal effect. We specify `y_type = "count"` and provide priors for the dispersion parameter $\phi$.

#### Prior Specification

We use `side = 2` (Two-sided noncompliance) and `ER = 1` (Exclusion Restriction assumed).

  * **Outcomes:** We estimate 4 models (Complier-Control, Complier-Treated, Never-taker, Always-taker).
  * **Phi:** We use a Gamma prior for the dispersion parameter.

<!-- end list -->

```{r fit_biva}
# Setup Priors
K_y <- 4 # 4 Outcome models
P_y <- 2 # Intercept + X

# Weakly informative priors for coefficients
beta_mean_y <- matrix(0, K_y, P_y)
beta_mean_y[, 1] <- 2.0 # Intercept prior (log scale)
beta_sd_y <- matrix(2, K_y, P_y)

# Priors for Strata selection
beta_mean_s <- matrix(0, 2, 2)
beta_sd_s <- matrix(1, 2, 2)

# Priors for Phi (Gamma shape/rate). Mean = shape/rate = 5/1 = 5.
phi_shape <- rep(5, K_y)
phi_rate <- rep(1, K_y)

# Initialize and Fit
ivobj <- biva$new(
  data = df,
  y = "Y", d = "D", z = "Z",
  x_ymodel = c("X"),
  x_smodel = c("X"),
  y_type = "count",          # <--- Key Argument
  ER = 1,
  side = 2,
  beta_mean_ymodel = beta_mean_y,
  beta_sd_ymodel = beta_sd_y,
  beta_mean_smodel = beta_mean_s,
  beta_sd_smodel = beta_sd_s,
  phi_shape_ymodel = phi_shape,
  phi_rate_ymodel = phi_rate,
  fit = TRUE,
  iter = 2000, chains = 4, seed = 2025
)
```

### Diagnostics

We check the trace plots to ensure the MCMC chains have converged for the mean outcome parameters.

```{r trace_plot, fig.align='center'}
ivobj$tracePlot()
```

We also check for weak instruments, which can inflate variance in IV models.

```{r weak_iv}
ivobj$weakIVTest()
```

### Results and Causal Estimates

We can now extract the **CACE** (Complier Average Causal Effect).

**Note:** In the context of count data, BIVA calculates the CACE as the **difference in expected counts** ($E[Y(1)] - E[Y(0)]$) for the complier population. This provides a highly interpretable metric (e.g., "The feature caused 4.5 extra logins").

```{r results}
# Strata Probabilities
ivobj$strataProbEstimate()

# Point Estimate (Median CACE)
cat("Estimated CACE (Median difference in counts): ", round(ivobj$pointEstimate(median = TRUE), 2))

# Credible Interval (95%)
ivobj$credibleInterval(width = 0.95)

# Probability of Positive Effect
ivobj$calcProb(a = 0)
```

The BIVA model successfully recovers a CACE close to the true simulated effect (approx 4.8), correcting for the bias seen in the naive analysis.

### Subgroup Prediction

A powerful feature of BIVA is estimating effects for specific user subgroups. For example, we can estimate the compliance probability and causal effect specifically for users with a high value of covariate $X$ (e.g., highly active users).

```{r prediction}
# Create a subgroup of users with X > 1
high_activity_users <- df[df$X > 1, ]

# Predict
ivobj$predict(new_data = high_activity_users, name = "High_Activity")

# Summarize
print(ivobj$predSummary(name = "High_Activity"))
```

This summary tells us the specific proportion of compliers within this high-activity group and the expected lift in logins for them.

-----

#### Conclusion

BIVA's extension to count outcomes allows for robust causal inference in settings with discrete, over-dispersed data. By modeling the data generating process explicitly with Negative Binomial distributions, we avoid the pitfalls of linear approximations and obtain interpretable results on the natural count scale.

---

#### BIVA

Licensed under the Apache License, Version 2.0.\
